---
title: "drugUse"
author: "Megan Nguyen"
date: "9/17/2018"
output: html_document
---
```{r}
library(tidyverse)
library(ROCR)
library(tree)
library(maptree)
library(class)
library(lattice)
```
Analyzing drug use

The first half of this assignment involves the analysis of drug use. The data set includes a total of 1885 observations on 32 variables. For each observation, 12 attributes are known:

• ID: number of record in original database. Used for reference only.

• Age: Age of the participant

• Gender: Gender of the participant (M/F)

• Education: Level of education of the participant

• Country: Country of current residence of the participant

• Ethnicity: Ethnicity of the participant

Many of the covariates have been transformed: some ordinal or categorical variables have been given numeric codes. Part of this problem will involve appropriately re-transforming these variables. The data also contains the following personality measurements:

• Nscore: NEO- FFI- R Neuroticism (Ranging from 12 to 60)

• Escore: NEO- FFI- R Extraversion (Ranging from 16 to 59)

• Oscore: NEO- FFI- R Openness (Ranging from 24 to 60)

• Ascore: NEO- FFI- R Agreeableness (Ranging from 12 to 60)

• Cscore: NEO- FFI- R Conscientiousness (Ranging from 17 to 59)

• Impulsive: Impulsiveness measured by BIS- 11

• SS: Sensation Seeking measured by ImpSS

Finally, participants were questioned concerning their use of 18 legal and illegal drugs (alcohol, amphetamines, amyl nitrite, benzodiazepine, cannabis, chocolate, cocaine, caffeine, crack, ecstasy, heroin, ketamine, legal highs, LSD, methadone, mushrooms, nicotine and volatile substance abuse) and one fictitious drug (Semeron) which was introduced to identify over-claimers. All of the drugs use the class system of CL0-CL6: CL0 = “Never Used”, CL1 = “Used over a decade ago”, CL2 = “Used in last decade”, CL3 = “Used in last year”, CL4 = “Used in last month”, CL5 = “Used in last week”, CL6 = “Used in last day”.

```{r}
drug_use <- read_csv('drug.csv',
col_names = c('ID','Age','Gender','Education','Country','Ethnicity',
'Nscore','Escore','Oscore','Ascore','Cscore','Impulsive',
'SS','Alcohol','Amphet','Amyl','Benzos','Caff','Cannabis',
'Choc','Coke','Crack','Ecstasy','Heroin','Ketamine',
'Legalh','LSD','Meth','Mushrooms','Nicotine','Semer','VSA'))
```

1. Logistic regression for drug use prediction

The data uses some strange encodings for variables. For instance, the gender variable has typedouble. Here the value -0.48246 means male and 0.48246 means female. Age was recorded at a set of categories but rescaled to a mean 0 numeric variable (we will leave that variable as is). Similarly education is a scaled numeric quantity (we will also leave this variable as is). We will however, start by transforming gender, ethnicity, and country to factors, and the drug response variables as ordered factors:

```{r}
drug_use <- drug_use %>% mutate_at(as.ordered, .vars=vars(Alcohol:VSA))
drug_use <- drug_use %>%
mutate(Gender = factor(Gender, labels=c("Male", "Female"))) %>%
mutate(Ethnicity = factor(Ethnicity, labels=c("Black", "Asian", "White",
"Mixed:White/Black", "Other",
"Mixed:White/Asian",
"Mixed:Black/Asian"))) %>%
mutate(Country = factor(Country, labels=c("Australia", "Canada", "New Zealand",
"Other", "Ireland", "UK", "USA")))
```

(a). We define a new factor response variable recent_cannabis_use which is “Yes” if a person has used cannabis
within a year, and “No” otherwise. This can be done by checking if the Cannabis variable is greater than
or equal to CL3. 

```{r}
drug_use <- drug_use %>%
  mutate(recent_cannabis_use = factor(ifelse(Cannabis > "CL2", "Yes", "No"), levels = c("No", "Yes")))
```

(b). We will create a new tibble that includes a subset of the original variables. We will focus on all variables
between age and SS as well as the new factor related to recent cannabis use. 

```{r}
drug_use_subset <- drug_use %>% select(Age:SS, recent_cannabis_use)
```
We split drug_use_subset into a training data set and a test data set called drug_use_train and drug_use_test.
The training data includes 1500 randomly sampled observation and the test data includes the
remaining observations in drug_use_subset.

```{r}
train <- sample(1:nrow(drug_use_subset), 1500)
drug_use_train <- drug_use_subset[train,]
drug_use_test <- drug_use_subset[-train,]
dim(drug_use_train)
dim(drug_use_test)
```

(c). We fit a logistic regression to model recent_cannabis_use as a function of all other predictors in
drug_use_train using the training data only. 

```{r}
glm_drug <- glm(recent_cannabis_use ~ ., data = drug_use_train, family = binomial)
summary(glm_drug)
```

2. Decision tree models of drug use

We construct a decision tree to predict recent_cannabis_use using all other predictors in drug_use_train.

```{r}
tree_parameters = tree.control(nobs=nrow(drug_use_train), minsize=10, mindev=1e-3)
drug_tree <- tree(recent_cannabis_use ~ ., data = drug_use_train, control = tree_parameters)
```

(a). Here we use 10-fold CV to select the a tree which minimizes the cross-validation misclassification rate.

```{r}
drug_tree_cv <- cv.tree(drug_tree, FUN = prune.misclass, K = 10)
min_dev <- min(drug_tree_cv$dev)
min_cv <- which(drug_tree_cv$dev == min_dev)
best_cv <- drug_tree_cv$size[min_cv]
best_cv
best_size <- min(best_cv)
best_size
```
(b). We prune the tree to the size found in the previous part and plot the tree.

```{r}
prune_drug_tree <- prune.tree(drug_tree, best = 7)
draw.tree(prune_drug_tree, nodeinfo = TRUE)
```
Country is split first in this decision tree.

(c). Confusion matrix for the test data and calculate the true positive rate (TPR) and false positive rate (FPR) for the confusion matrix. 
```{r}
drug_tree_pred <- predict(prune_drug_tree, drug_use_test, type = "class")
drug_tree_pred
true_drug <- drug_use_test$recent_cannabis_use
error <- table(drug_tree_pred, true_drug)
error
```
```{r}
TPR <- error[2,2]/(error[2,2]+error[1,2])
FPR <- error[2,1]/(error[2,1]+error[1,1])
TPR
FPR
```

3. Model Comparison

(a). ROC curves for both the logistic regression fit and the decision tree

```{r}
#GLM Predictions
glm_test <- glm(recent_cannabis_use~., drug_use_test, family = binomial)
glm_pred_test <- predict(glm_test, type = "response")
true_test <- drug_use_test$recent_cannabis_use
glm_prediction <- prediction(glm_pred_test, true_test)
glm_perf <- performance(glm_prediction, measure = "tpr", x.measure = "fpr")

#Generate Tree Predictions
drug_tree_pred
drug_tree_pred <- data.frame(drug_tree_pred)
#Prediction type
tree_prediction <- prediction(drug_tree_pred, c(true_test))
#Generate tree performance
tree_perf = performance(tree_prediction, measure="tpr", x.measure="fpr")

plot(tree_perf, col = 2, lwd = 3, main = "ROC Curve Tree")
abline(0,1)
par(new = TRUE)
plot(glm_perf, axes = FALSE, col = 3, lwd = 3, xlab = "", ylab = "")
```
(b). AUC for both models and print them. 

```{r}
#AUC
glm_auc <- performance(glm_prediction, "auc")@y.values
glm_auc

tree_auc <- performance(tree_prediction, "auc")@y.values
tree_auc
```
The glm has a greater AUC

4. Clustering and dimension reduction for gene expression data

This problem involves the analysis of gene expression data from 327 subjects from Yeoh et al (2002). The data set includes abundance levels for 3141 genes and a class label indicating one of 7 leukemia subtypes the patient was diagnosed with.

```{r}
leukemia_data <- read_csv("leukemia_data.csv")
```

(a). The class of the first column of leukemia_data, Type, is set to character by default, so we convert the Type column to a factor. 
```{r}
leukemia_data <- leukemia_data %>%
  mutate(Type = factor(Type))
type_table <- table(leukemia_data$Type)
type_table
```
The BCR-ABL subtype occurs the least.

(b). We run PCA on the leukemia data and plot the proportion of variance explained by each principal component (PVE) and the cumulative PVE side-by-side.

```{r}
pc <- prcomp(leukemia_data[,-1], center = TRUE, scale = TRUE)
pc
```

```{r}
sdev <- pc$sdev
pve <- sdev^2 / sum(sdev^2)
cumulative_pve <- cumsum(pve)
## This will put the next two plots side by side
par(mfrow=c(1, 2))
## Plot proportion of variance explained
plot(pve, type="l", lwd=3)
plot(cumulative_pve, type="l", lwd=3)
```

(c). We use the results of PCA to project the data into the first two principal component dimensions. prcomp returns this dimension reduced data in the first columns of x. We plot the data as a scatter plot

```{r}
rainbow_colors <- rainbow(7)
plot_colors <- rainbow_colors[leukemia_data$Type]

plot(pc$x, col = plot_colors, cex = 0)
text(pc$x, labels = leukemia_data$Type, col = plot_colors, cex = 0.30)
```

T-ALL is separated from the other groups along the PC1 axis.
```{r}
pc1 <- abs(pc$x[,1])
pc1_sort <- sort(pc1, decreasing = TRUE)
head(pc1_sort)
```


