---
title: "drugUse"
author: "Megan Nguyen"
date: "9/17/2018"
output: html_document
---
```{r}
library(tidyverse)
library(ROCR)
library(tree)
library(maptree)
library(class)
library(lattice)
```
Analyzing drug use
The first half of this homework involves the analysis of drug use. The data set includes a total of 1885
observations on 32 variables. A detailed description of the data set can be found here. For each observation,
12 attributes are known:
• ID: number of record in original database. Used for reference only.
• Age: Age of the participant
• Gender: Gender of the participant (M/F)
• Education: Level of education of the participant
• Country: Country of current residence of the participant
• Ethnicity: Ethnicity of the participant
Many of the covariates have been transformed: some ordinal or categorical variables have been given numeric
codes. Part of this problem will involve appropriately re-transforming these variables. The data also contains
the following personality measurements:
• Nscore: NEO- FFI- R Neuroticism (Ranging from 12 to 60)
• Escore: NEO- FFI- R Extraversion (Ranging from 16 to 59)
• Oscore: NEO- FFI- R Openness (Ranging from 24 to 60)
• Ascore: NEO- FFI- R Agreeableness (Ranging from 12 to 60)
• Cscore: NEO- FFI- R Conscientiousness (Ranging from 17 to 59)
• Impulsive: Impulsiveness measured by BIS- 11
• SS: Sensation Seeking measured by ImpSS
Finally, participants were questioned concerning their use of 18 legal and illegal drugs (alcohol, amphetamines,
amyl nitrite, benzodiazepine, cannabis, chocolate, cocaine, caffeine, crack, ecstasy, heroin, ketamine,
legal highs, LSD, methadone, mushrooms, nicotine and volatile substance abuse) and one fictitious
drug (Semeron) which was introduced to identify over-claimers. All of the drugs use the class system
of CL0-CL6: CL0 = “Never Used”, CL1 = “Used over a decade ago”, CL2 = “Used in last decade”,
CL3 = “Used in last year”, CL4 = “Used in last month”, CL5 = “Used in last week”, CL6 = “Used in last day”.
```{r}
drug_use <- read_csv('drug.csv',
col_names = c('ID','Age','Gender','Education','Country','Ethnicity',
'Nscore','Escore','Oscore','Ascore','Cscore','Impulsive',
'SS','Alcohol','Amphet','Amyl','Benzos','Caff','Cannabis',
'Choc','Coke','Crack','Ecstasy','Heroin','Ketamine',
'Legalh','LSD','Meth','Mushrooms','Nicotine','Semer','VSA'))
```

1. Logistic regression for drug use prediction

This problem has 4 parts for 131 students and 5 parts for 231 students. As mentioned, the data uses
some strange encodings for variables. For instance, you may notice that the gender variable has type
double. Here the value -0.48246 means male and 0.48246 means female. Age was recorded at a set
of categories but rescaled to a mean 0 numeric variable (we will leave that variable as is). Similarly
education is a scaled numeric quantity (we will also leave this variable as is). We will however, start by
transforming gender, ethnicity, and country to factors, and the drug response variables as ordered factors:
```{r}
drug_use <- drug_use %>% mutate_at(as.ordered, .vars=vars(Alcohol:VSA))
drug_use <- drug_use %>%
mutate(Gender = factor(Gender, labels=c("Male", "Female"))) %>%
mutate(Ethnicity = factor(Ethnicity, labels=c("Black", "Asian", "White",
"Mixed:White/Black", "Other",
"Mixed:White/Asian",
"Mixed:Black/Asian"))) %>%
mutate(Country = factor(Country, labels=c("Australia", "Canada", "New Zealand",
"Other", "Ireland", "UK", "USA")))
```

(a). Define a new factor response variable recent_cannabis_use which is “Yes” if a person has used cannabis
within a year, and “No” otherwise. This can be done by checking if the Cannabis variable is greater than
or equal to CL3. Hint: use mutate with the ifelse command. When creating the new factor set levels
argument to levels=c("No", "Yes") (in that order).
```{r}
drug_use <- drug_use %>%
  mutate(recent_cannabis_use = factor(ifelse(Cannabis > "CL2", "Yes", "No"), levels = c("No", "Yes")))
```

(b). We will create a new tibble that includes a subset of the original variables. We will focus on all variables
between age and SS as well as the new factor related to recent cannabis use. Create drug_use_subset with
the command:
```{r}
drug_use_subset <- drug_use %>% select(Age:SS, recent_cannabis_use)
```
Split drug_use_subset into a training data set and a test data set called drug_use_train and drug_use_test.
The training data should include 1500 randomly sampled observation and the test data should include the
remaining observations in drug_use_subset. Verify that the data sets are of the right size by printing
dim(drug_use_train) and dim(drug_use_test).
```{r}
train <- sample(1:nrow(drug_use_subset), 1500)
drug_use_train <- drug_use_subset[train,]
drug_use_test <- drug_use_subset[-train,]
dim(drug_use_train)
dim(drug_use_test)
```

(c). Fit a logistic regression to model recent_cannabis_use as a function of all other predictors in
drug_use_train. Fit this regression using the training data only. Display the results by calling the summary
function on the logistic regression object.
```{r}
glm_drug <- glm(recent_cannabis_use ~ ., data = drug_use_train, family = binomial)
summary(glm_drug)
```

2. Decision tree models of drug use

This problem has 3 parts for all students (25 points total)
Construct a decision tree to predict recent_cannabis_use using all other predictors in drug_use_train.
Set the value of the argument control = tree_parameters where tree_parameters are:
tree_parameters = tree.control(nobs=nrow(drug_use_train), minsize=10, mindev=1e-3)
This sets the smallest number of allowed observations in each leaf node to 10 and requires a deviance of at
least 1e-3 to split a node.
```{r}
tree_parameters = tree.control(nobs=nrow(drug_use_train), minsize=10, mindev=1e-3)
drug_tree <- tree(recent_cannabis_use ~ ., data = drug_use_train, control = tree_parameters)
```

(a). Use 10-fold CV to select the a tree which minimizes the cross-validation misclassification rate.
Use the function cv.tree, and set the argument FUN=prune.misclass. Note: you do not need to
use a do.chunk function since the tree package will do cross validation for you. Find the size
of the tree which minimizes the cross validation error. If multiple trees have the same minimum
cross validated misclassification rate, set best_size to the smallest tree size with that minimum rate.
```{r}
drug_tree_cv <- cv.tree(drug_tree, FUN = prune.misclass, K = 10)
min_dev <- min(drug_tree_cv$dev)
min_cv <- which(drug_tree_cv$dev == min_dev)
best_cv <- drug_tree_cv$size[min_cv]
best_cv
best_size <- min(best_cv)
best_size
```
(b). Prune the tree to the size found in the previous part and plot the tree using the draw.tree function
from the maptree package. Set nodeinfo=TRUE. Which variable is split first in this decision tree?
```{r}
prune_drug_tree <- prune.tree(drug_tree, best = 7)
draw.tree(prune_drug_tree, nodeinfo = TRUE)
```
Country is split first in this decision tree.

(c). Compute and print the confusion matrix for the test data using the function table(truth,
predictions) where truth and predictions are the true classes and the predicted classes from the tree
model respectively. Note: when generated the predicted classes for the test data, set type="class" in the
predict function. Calculate the true positive rate (TPR) and false positive rate (FPR) for the confusion
matrix. Show how you arrived at your answer.
```{r}
drug_tree_pred <- predict(prune_drug_tree, drug_use_test, type = "class")
drug_tree_pred
true_drug <- drug_use_test$recent_cannabis_use
error <- table(drug_tree_pred, true_drug)
error
```
```{r}
TPR <- error[2,2]/(error[2,2]+error[1,2])
FPR <- error[2,1]/(error[2,1]+error[1,1])
TPR
FPR
```

3. Model Comparison

This problem has 2 parts for all students worth 15 points total.
(a). Plot the ROC curves for both the logistic regression fit and the decision tree on the same plot. Use
drug_use_test to compute the ROC curves for both the logistic regression model and the best pruned tree
model.
```{r}
#GLM Predictions
glm_test <- glm(recent_cannabis_use~., drug_use_test, family = binomial)
glm_pred_test <- predict(glm_test, type = "response")
true_test <- drug_use_test$recent_cannabis_use
glm_prediction <- prediction(glm_pred_test, true_test)
glm_perf <- performance(glm_prediction, measure = "tpr", x.measure = "fpr")



#Generate Tree Predictions
drug_tree_pred
drug_tree_pred <- data.frame(drug_tree_pred)
#Prediction type
tree_prediction <- prediction(drug_tree_pred, c(true_test))
#Generate tree performance
tree_perf = performance(tree_prediction, measure="tpr", x.measure="fpr")

plot(tree_perf, col = 2, lwd = 3, main = "ROC Curve Tree")
abline(0,1)
par(new = TRUE)
plot(glm_perf, axes = FALSE, col = 3, lwd = 3, xlab = "", ylab = "")
```
(b). Compute the AUC for both models and print them. Which model has larger AUC?
```{r}
#AUC
glm_auc <- performance(glm_prediction, "auc")@y.values
glm_auc

tree_auc <- performance(tree_prediction, "auc")@y.values
tree_auc
```
The glm has a greater AUC

4. Clustering and dimension reduction for gene expression data
This problem involves the analysis of gene expression data from 327 subjects from Yeoh et al (2002). The
data set includes abundance levels for 3141 genes and a class label indicating one of 7 leukemia subtypes
the patient was diagnosed with. The paper describing their analysis of this data can be found here. Read
in the csv data in leukemia_data.csv. It is posted on Piazza in the resources tab with the homework:
```{r}
leukemia_data <- read_csv("leukemia_data.csv")
```
This problem has 4 parts for 131 students for a total of 35 points and 7 parts for 231 students for a total of
50 points.
(a). The class of the first column of leukemia_data, Type, is set to character by default. Convert the
Type column to a factor using the mutate function. Use the table command to print the number of patients
with each leukemia subtype. Which leukemia subtype occurs the least in this data?
```{r}
leukemia_data <- leukemia_data %>%
  mutate(Type = factor(Type))
type_table <- table(leukemia_data$Type)
type_table
```
The BCR-ABL subtype occurs the least.

(b). Run PCA on the leukemia data using prcomp function with scale=TRUE and center=TRUE
(this scales each gene to have mean 0 and variance 1). Make sure you exclude the Type column
when you run the PCA function (we are only interested in reducing the dimension of the
gene expression values and PCA doesn’t work with categorical data anyway). Plot the proportion
of variance explained by each principal component (PVE) and the cumulative PVE side-by-side.
```{r}
pc <- prcomp(leukemia_data[,-1], center = TRUE, scale = TRUE)
pc
```

```{r}
sdev <- pc$sdev
pve <- sdev^2 / sum(sdev^2)
cumulative_pve <- cumsum(pve)
## This will put the next two plots side by side
par(mfrow=c(1, 2))
## Plot proportion of variance explained
plot(pve, type="l", lwd=3)
plot(cumulative_pve, type="l", lwd=3)
```

(c). Use the results of PCA to project the data into the first two principal component dimensions. prcomp
returns this dimension reduced data in the first columns of x. Plot the data as a scatter plot using plot
function with col=plot_colors where plot_colors is defined.
This will color the points according to the leukemia subtype. Add the leukemia type labels to the plot using
text with labels argument set to the leukemia type and the col to plot_colors (it may help legibility to
make the points on the plot very small by setting cex to a small number). Which group is most clearly
separated from the others along the PC1 axis? Which genes have the highest absolute loadings for PC1 (the
genes that have the largest weights in the weighted average used to create the new variable PC1)? You can
find these by taking the absolute values of the first principal component loadings and sorting them. Print the
first 6 genes in this sorted vector using the head function.
```{r}
rainbow_colors <- rainbow(7)
plot_colors <- rainbow_colors[leukemia_data$Type]

plot(pc$x, col = plot_colors, cex = 0)
text(pc$x, labels = leukemia_data$Type, col = plot_colors, cex = 0.30)
```

T-ALL is separated from the other groups along the PC1 axis.
```{r}
pc1 <- abs(pc$x[,1])
pc1_sort <- sort(pc1, decreasing = TRUE)
head(pc1_sort)
```


